\hypertarget{quantum-depletion}{%
\section{Quantum depletion}\label{quantum-depletion}}

\hypertarget{introduction}{%
\subsection{Introduction}\label{introduction}}

\begin{verbatim}
A Bose-Einstein condensate (BEC) forms when a gas of bosons is cooled
below a critical temperature, where the majority of atoms occupy the
ground state of the system. At non-zero temperature, the condensate will
be depleted by thermal excitations, where some atoms occupy higher
energy states with occupation probabilities following a Bose-Einstein
statistics. However, even at absolute zero, the
condensate will still be depleted by quantum fluctuations of the quasiparticle vacuum, a phenomenon known as quantum depletion. The spectrum of fluctuations in the
quasiparticle vacuum gives rise to an asymptotic momentum distribution
that follows a $k^{-4}$ power law[Qu2016].

A recent experiment [Chang2016] was able to see evidence of quantum
depletion in the large $k$ tails of an interacting BEC released from an
optical dipole trap, using the high-resolution and large dynamic range
of detection offered by ultracold metastable helium atoms (He*)
[Vassen2012]. A separate experiment on a strongly interacting BEC in a
uniform potential probed with Bragg scattering showed that the depletion
could be tuned with the interaction strength [Lopes2017].

According to current theory however, the quantum depletion seen in
[Chang2016] is not expected to be observed, as the interactions between
atoms should cause the $k^{-4}$ tails to adiabatically vanish during the
expansion after trap switch off [Qu2016]. During the expansion, which
is taken to be hydrodynamic, the contact parameter tends to zero, which
eliminates the depletion tail for sufficiently long expansions. This
viewpoint was supported by an earlier experimental result that failed to
observe the tails [Makotyn2014]. However, it is unclear whether the
high momentum tails, which are by definition fast moving, are in fact
accurately described by this model or if their shorter interaction time
with the high density region of the BEC (in a semi-classical picture)
means they do in fact survive the expansion. To help resolve this
discrepancy, we have conducted a similar experiment to [Chang2016]
looking at the momentum tails of a BEC of He\* released from a magnetic
trap and detected after a long time of flight.
\end{verbatim}

\hypertarget{methods}{%
\subsection{Methods}\label{methods}}

\begin{verbatim}
Our experiments start with a condensate of helium atoms in the
$M_J = +1$ state of the metastable $2^3 S_1$ manifold in a BiQUIC
magnetic trap [Dall2007]. We use two trap configurations with characteristic frequencies $\sim 2\pi\cdot (312,312,52)$ and $2\pi\cdot (666,666,69)$ Hz. The trap is switched off in $\sim 100\mu s$ and the atoms allowed to fall 848mm, where they are detected with a
multi-channel plate and delay line detector [Manning2010]. To avoid distortion of the condensate by stray magnetic fields, we transferred approximately 25% of the condensate into the magnetically insensitive $M_J=0$ state by applying a chirped sine wave 2ms after trap switch-off, swept from 1.6 to 2.6 MHz in 1ms, while a DC magnetic field remained on to preserve the Zeeman splitting between $M_J = \{0,\pm 1\}$ states. A magnetic field gradient is switched on after the state transfer pulse, which ensures the magnetically sensitive $M_J = \pm 1$ states are deflected enough to miss the detector. We varied total atom number between X and $Y\times 10^5$ atoms per condensate by varying the endpoint of the evaporative cooling ramp. 

After release from the trap the atoms expand ballistically into the
far-field regime, where their position on the detector corresponds to
their momentum at trap release. Therefore we can reconstruct the far-field
momentum distribution $n^*(\textbf{k})$ from our detector data. The
observed momentum profile $n^*(\textbf{k})$ cannot be identified
completely with the in-trap momentum distribution $n_0(\textbf{k})$ as
repulsive interatomic forces distort the profile immediately after
release from the trap. Structure is visible over five orders of
magnitude in density, comprised of the condensed, thermal, and depleted
fraction before the signal drops below the dark-count background noise.
There is some saturation evident around the low momentum values due to
the high atom flux during BEC impact. Our analysis does not include this part of the momentum profile, and we find evidence that saturation is only an issue during the peak flux, not in the thermal or depleted fractions.

Since for these experiments a precise knowledge of the total atom number and trap frequency is important, we measure the trapped population and trapping frequency at regular intervals throughout each experimental run, using a procedure that avoids detector saturation to measure the atom number and precise determination of the trap frequency. We identified a contamination of the depletion signal by stray counts attributable to the m=+1 state, and calibrate for this by including shots without the RF transfer step. We also apply corrections for the detector dark count rate and an increased background rate on the leading edge of the condensate (details about the calibration are included in the supplementary materials)

The Tan contact parameter is extracted by a fit to BEC density profile after subtracting the calibrated noise profiles. The functional form includes contributions from the thermal fraction and depleted fraction and includes the free parameters $N_{th}$ the number of thermal atoms, $\lambda_{dB} = h/\sqrt{2\pi m k_B T}$ the thermal de Broglie wavelength, and $C$ is the scale of the fitted power law.

FIGS
Experimental sequence
\end{verbatim}

\hypertarget{results}{%
\subsection{Results}\label{results}}

\begin{verbatim}
FIGS:
Condensate density profile & scaled, with fit
Contact vs Density plot
\end{verbatim}

According to theoretical predictions based on the local density
approximation {[}Chang2016{]}, the contact constant \(C_{LDA}\) of the
depleted fraction should vary linearly with the peak condensate density
\(n_0\). This was found in previous experiments{[}Chang2016{]}, although
their scaling factor greater by a factor of \$\sim\$6.5. In Fig. XX, we
plot our measured values of \(C\) and \(n_0\), identifying the fit
parameter \(C\) with the contact constant \(C_{LDA}\). A linear fit to
the contact constant variation finds a gradient approximately 1.7 times
more than the theory predicts, closer to theory than the previous
measurement

There are some possible explanations for the disagreement between LDA
theory and experiment. This experiment requires resolution of a strong
signal over nearly five orders of magnitude, careful calibration, and
accurate parameter estimation with sparse and noisy data, and power-law
fits are notoriously difficult.

The mapping of the in-trap to free-particle momentum distribution for a
fast (but not instantaneous) trap release in 3D has no analytic solution
and is challenging to accurately model numerically. Indeed, switch-off
timescale may be relevant (compare {[}Chang2016{]} with {[}Qu2016{]}).
The RF pulse may increase the Penning ionization rate in the depolarized
sample, even though the density is low, which would alter the momentum
profile. While applying the magnetic separation pulse, the condensates
may scatter off each other.

\hypertarget{discussion}{%
\subsection{Discussion}\label{discussion}}

This suggests that the originally observed result is indeed a real
physical effect rather than experimental error. Hopefully our result
will stimulate new theoretical interest in developing an explanation of
why the quantum depletion tails can be observed after an interacting
expansion, as well as try to resolve the discrepancy with theory.

A possible future experimental extension is to outcouple atoms using a
broadened Raman transition. This would have the benefit of moving the
outcoupled atoms through the cloud rapidly, minimising the effects of
repulsive inter-atomic interactions that normally affect the outcoupled
profiles. This should shed some light on whether interactions during
trap switch off are the origin of the momentum tails. One could test the
importance of particle scattering or Penning ionization by delaying the
Rabi transfer to the \(M_J = 0\) state, but this could increase
vulnerability to stray magnetic fields. There is scope to optimize the
RF transfer sequence to achieve much higher transfer efficiency for a
stronger signal. \#\# Issues\\
Sys: Background Stat: Fitting power laws, weak signals \#\# Method\\
\# Supplementary materials

FIGS Combined time profiles Thermal transfer calibration Interleaving
sequence - main text?

\hypertarget{calibration}{%
\subsection{Calibration}\label{calibration}}

We assume that the RF pulse and magnetic separation of the condensates
does not affect their far-field distribution. This may be false if the
inter-condensate scattering rate is high (calculation/reference
pending). The RF pulse transfers atoms into the different(mJ)states with
efficiency(\etaJ), therefore we estimate the momentum density of the
initial condensate via

{[}n(k) = \frac{1}{\eta0}\left(\bar\{n\}0 - \eta1\bar\{n\}1 -
1+\eta1)(\delta+\lambda)\right){]}

where (\bar\{n\}\{1\}) refer respectively to the count densities
obtained from the measurement and spin mixing calibration shots. We
produce a background calibration for((\delta+\lambda))by defining
background density in the upper and lower hemisphere
of(\bar\{\delta\}\pm)to be the backgrounds obtained from the dark count
and leak calibrations, respectively. We determined the transfer
efficiencies(\etaJ)from the initially released condensate to the
other(mJ)states by separating the clouds with a magnetic field gradient
generated by auxiliary field coils. (Refinement pending). We calibrate
the dark count rate by building a spherical histogram centred at the
same position on the detector as the(mJ=0)condensate, but 2 seconds
after the condensate impact and before the trap loading sequence begins,
and calibrate the trap leak similarly but centering the histogram about
a point earlier in time than the BEC impact. We calibrate the
contamination by stray(mJ=1)counts by running the depletion measurement
sequence without the RF transfer, obtaining the distribution of stray
counts, plus the background count rate. By subtracting our empirical
measurement of the background, we obtain the estimated density of the
stray counts,(n1 = \bar\{n\}1 - (\delta+\lambda)).

The total atom number(N)and trapping frequencies(\omegai)are determined
by a pulsed atom-laser measurement, which we describe in detail in
{[}another publication{]}. A fit of the empirical thermal fraction
produces a measurement of the temperature, through which we may
determine the condensed number(N0).

Average dark count rate is determined to be(5.6228E3 Hz/m\^{}2),
or(9.3713E-16 (m\textsuperscript{\{-1\})}\{-3\}). The latter is
calculated by converting 1 sec interval into a distance c.f. the centre
of mass velocity of the BEC - about 4m or 4E6 micron. Sequence \&
Calibration stages Total number

\begin{verbatim}
    Well this isn't anything new, I just use the pulsed atom laser in the same way as other chapters. Because I care more about number here I'd like to spend a bit of time thinknig about number errors that could crop up here which are perhaps less important to treat than they are in tuneout, which will be a busy chapter in its own right. 
Raw profiles

    Centering algorithm to align BECs and compile to single shot.
    Create histogram in spherical coordinates.
    Compile and apply calibration corrections. 
    Do some fancy fitting - theory-driven pre guesses
    T from fits and consistnency 
Trap leakage

    Trap leakage is surely number-dependent. Who knows what the law actually is - there is probably some combination of majorana and ??? - perhaps could ballpark what the leak rate should be, see if it squares with what I observe. But yeah it results in an increased flux rate, you can see it in the time trace. The best way to deal with it would be to extend the hold for ages, and then get heaps of statistics, but in the end I wound up using the 20-odd ms window of hold before drop in each case. This still gets to within ??? some margin of error when using all the data I have for a given run. Given the small count rate it isn;t at all obvious that I would be able to do this shot-for-shot; not sure what type of noise this is. 
Background

    This one was easy. I just took the dark count rate from everywhere else; make a histogram displaced on the detector, assuming the dark counts and hotspots are time invariant. Are they? Worth a look I guess. But this is just like camera dark-fielding. That could be a great analogy actually! There are three channels, some background, and, well, the leakage but whatever let's sort of ignore that. Let's show it's Poissonian, maybe? Just to show that it really is uncorrelated events?
Spin mixing

    Well, this one was OK, I just ran the experiment without the transfer. It shows up fine, I used the center found in the AVERAGE (I think) which might be worth looking at - should it be better to use the centre fixed from the proximate shots? idk. Depends on whether there's genuine drift or just jitter in BEC arrival position. Lol, I guess one could do an allan variance of this one also, that would tell you the best period to average across. Actually, yeah that is kind of cool. Perhaps including that much detail in the 'possible improvements' part. The origin of this isn't clear, hey. It is unlikely to be m=1 atoms, honestly, as they would probably be movable with the Z coil after the 'spray'. The spray moment suggests it's something in the fall path that seems to be an issue - one can control a little bit where it impacts by adjusting push coil timigns - but never quite entirely get rid of it. There is also the question of the m=-1 atoms floating around. But then, calibrating for that would involve using a different RF sequence, perhaps, and at the end of the day maybe the best thing to do would be to find a Rabi transfer setup that pushed everything into M=0 anyhow, and see whether the symmetry is a problem there. Something I'd like to do is to drop an entire BEC, center it very carefully and then reflect it in 3D, and subtract histograms. But that's a mini project for another time and might not really demonstrate much. I would need to include at this point an argument to gauge the systematic error, for example by spreading the entire population uniformly or worst-case just in the power law.  Then see how much a difference that woudl make.
Transfer fraction calibration

    Verification that saturation is not a thing - also kinda pulls out the flux rate wehre saturation is a problem (could do this in 3D, obvs nonlinear but whatever, it's a ballpark)
    Verifies that m=0 halo is not distorted significantly in the transfer process
    Verifies that at least within the thermal part, no k-dependence of transfer
    Model of sweep - would it have been practical to get the max transfer without any momentum-dependences? - what is the doppler shift of the depleted atoms, anyhow? Not to mention the RF field is certainly not spatially coherent hey
    How accurately is this determined? 

    Processing

My general dream for these sections is a graph of the program or some kind of systems diagram, maybe nested screenshots or pseudocode would do the trick and then include the code as an appendix hey?

## Definition
\end{verbatim}

In the image obtained from the MCP-DLD detector of the shots in which
the quantum depletion is observed, the pointwise density of atom
detection events in momentum space can be described by

{[} n(\textbf{k}) = \sum{m=1}\^{}3 nm(\textbf{k},N,\omega{x,y,z},T) +
\delta(\textbf{k}) + \lambda(N0,\bar\{\omega\}\textbf{k})\Theta(-kz) {]}

The first terms(nm)refer to the detected density of atoms from
the(mJ=m)condensates, which depends on the momentum vector(\textbf{k})on
the total atom number(N), the temperature (T), and harmonic trapping
frequencies(\omega{x,y,z}). The dark count rate(\delta)is
momentum-dependent due to its non-uniformity on the detector. The
measured trap leak rate(\lambda)stops when the trap is released, hence
the Heaviside(\Theta)function ensures this term only contributes on the
lower side of the falling BEC. Below we adopt the
notation((\delta+\lambda) = \delta +
\lambda\Theta(\theta))where(\theta)is the elevation angle from the plane
of the detector subtending from the centre of the condensate. Details of
the measurement and calibration of each of these terms is discussed in
section calibration.

\hypertarget{contributions-to-the-model}{%
\subsection{Contributions to the
model}\label{contributions-to-the-model}}

The count density from the falling BECs depend explicitly on the trapped
population and on the trapping frequencies. This is most obviously true
for the condensate of atoms in the(mJ=0)state, and also for
the(mJ=\pm 1)condensates when all three condensates land undisturbed on
the detector during calibration of the transfer efficiency.

In the depletion detection shots, we have observed a remaining presence
of(mJ=1)atoms, the cause of which is unclear but is suspected to be due
to a collision process with a feature inside the chamber. Regardless of
the cause, we calibrate for this contamination (also called
\textit{spin mixing} elsewhere). The count density of the(mJ=1)states
depends on the trapping frequencies also - this could be because the
trap centre shifts with changing trap frequencies, changing the details
of the spin mixing process during freefall. We cannot rule out the
presence of(mJ=-1)on the detector also. Calibrating for this
contribution is less straightforward, as discussed below. The
temperature dependence of the condensate density is expected to manifest
only as a change in the thermal fraction - calculation of the thermal
depletion contributions are pending.

The contributions from the dark count rate are assumed to be
time-invariant and independent of any properties of the condensate
itself. There may be a change in the dark count rate while the detector
is subject to high atom fluxes, but these are not expected to be a
problem for the following reasons: One, we observe very similar thermal
tails both above and below the condensate, suggesting that, at least,
the quantum efficiency and dark count rates are not significantly
different. Two, although there are some temporary hotspots on the
detector during the peak BEC flux, these are only observed co-temporally
with the falling BEC. The quantum depletion is detected far beyond the
regions where this effect is noticable, and so they are not expected to
contaminate the signal.

The leak rate should be expected to be dependent on the BEC number. The
most simple model would be an exponential decay of BEC number with some
collection efficiency by the detector. There may also be
density-dependent collision effects feeding the trap leak rate driven by
changing the trapping frequencies, which could also alter any spatial
structure in the leak rate due to a shift in trap position.

\hypertarget{obtaining-the-tan-constant}{%
\subsection{Obtaining the Tan
constant}\label{obtaining-the-tan-constant}}

The far-field density of a harmonically trapped condensate is described
by contributions from the condensed, thermal, and depleted populations,

{[} n(\textbf{k}) = n\{BEC\}(\textbf{k}) + n\{T\}(\textbf{k}). {]}

The momentum density of the thermal fraction is {[}Pitaevskii \&
Stringari{]} {[} n\{T\}(\textbf{k}) =
\frac{1}{\left(\lambdaT m \bar{\omega}\right)^3}
g\{3/2\}\left(e\textsuperscript{\{-\beta k}2/2m\hbar\^{}2\}\right) {]}
where(T)is the temperature,(\lambdaT = h\^{}2/\sqrt{2\pi m kB T})is the
thermal de Broglie wavelength,(\bar\{\omega\})is the geometric mean of
the trapping frequencies,(\beta=1/ kB T)is the thermodynamic Beta,
and(g\{3/2\}(z))is the Bose integral {[} g\{3/2\}(z) =
\frac{2}{\sqrt{\pi}}\int0\textsuperscript{\infty \frac{\sqrt{x}}{z^{-1}e^x-1}=\sum{l=1}}\infty \frac{z^l}{l^{3/2}}.
{]}

The short-wavelength density of the BEC momentum density is described by
the Thomas-Fermi approximation. The asymptotic momentum distribution is
that of the quantum depletion, {[} n(\textbf{k}) =
\frac{C\infty}{(2\pi)^3 k^4}, {]}

where(C\infty)is the universal Tan constant, defined
by(\mathcal{C}{\infty}=\textrm{lim}\{\textbar{}\textbf{k}\textbar{}\to\infty\}(2\pi)\^{}3
\textbar{}\textbf{k}\textbar\^{}4 n(\textbf{k})), and for a harmonically
trapped Bose gas is(\frac{64\pi^2}{7}a\{s\}\^{}\{2\}N0 n0)in the local
density approximation {[}Chang et al{]}.

The asymptotic behaviour of the single-particle probability density
function for a particle in the ground state (not in the thermal
fraction) is therefore(\frac{64\pi^2}{7}a\{s\}\^{}\{2\} n0), hence the
contact constant can be seen as a parameter defining a probability
distribution, which we estimate using the procedure described in the
next section.
