To-do list:\newline
In \verb|{00_abstract.tex}|:
\begin{itemize}
\item {Write abstract}
\item {Ensure list is up to date}
\end{itemize}
In \verb|{00_frontmatter.tex}|:
\begin{itemize}
\item {Acknolwedgements}
\end{itemize}
In \verb|{10_overview.tex}|:
\begin{itemize}
\item {Make sensible; precis = thesis outline}
\item {Prologue = context/history of the field/where this thesis fits in}
\end{itemize}
In \verb|{11_introduction.tex}|:
\begin{itemize}
\item {Waiting for Andrew comments}
\end{itemize}
In \verb|{12_apparatus.tex}|:
\begin{itemize}
\item {These are wrong, will fix}
\end{itemize}
In \verb|{12_apparatus_test.tex}|:
\begin{itemize}
\item {Fix references; generally a big task. Standardize citation formats with py script?}
\item {Whenever a general discussion is presented, ensure it closes with a direct connection to the experiment with adequate specifications.}
\item {Properly caption this figure}
\item {check math. Include specific laser intensities/detunings for some parts of the experiment.}
\item {this; dig out zeeman code}
\item {Is it just a pair of coils? I thought there was one on each window...}
\item {how to get number from sat fluro}
\item {Define far-field.}
\item {Caption this figure and link with the text. Add pictures of detector elsewhere.}
\item {Maybe include a figure showing this dark rate?}
\item {So what parameters do we use for our atom lasers then?  All this background is fine, but in a chapter on the experiment you need some more details of what our system actually has}
\end{itemize}
In \verb|{13_lattice_build.tex}|:
\begin{itemize}
\item {Find the balance between digression (for motivation's sake) and cohesion with the Overview chapter}
\item {Heavy edits on motivation}
\item {Double-well system example? Wannier function and ED code?}
\item {Incorporate optical diagrams properly}
\item {Description of dipole optics}
\item {Diagram and samples of absorption imaging}
\item {Discussion and pictures of build}
\item {Use the line `In this section we deal with the nuts and bolts (literally) of assembling a cold-atom machine'}
\item {Tales of dipole alignment; abs img 'comet', the 'cleft' and prior misadventures, dipole beam profiling, reason lattice beam not locked}
\item {More to say/cite about correlation functions and the many body problem, callback to quasiparticles/emergent stuff...}
\end{itemize}
In \verb|{20_metrology.tex}|:
\begin{itemize}
\item {How state-of-the-art theory works, the contributions from experiments. Add citations. Look over the ol Pachucki papers etc for primary refs. }
\item {Some context about the history of precision spectroscopy - lift from later chapter? Put tuneout section in context, at least with a couple sentences, but approach distinctly later}
\item {orientation dependence goes somewhere in here (eg near the $5\triplet D$ section?)}
\end{itemize}
In \verb|{21_transitions.tex}|:
\begin{itemize}
\item {Shorter title or fix page headers}
\item {Put figures in the right place}
\end{itemize}
In \verb|{30_manybody.tex}|:
\begin{itemize}
\item {Possibly don't need this section. But could pare down significantly to give context. Just need to discuss the QD chapter.}
\item {How would you observe this?  What would it look like?}
\end{itemize}
In \verb|{31_depletion.tex}|:
\begin{itemize}
\item {Density plots out to larger $|k|$ to show descent into noise floor? How far could we push it until we hit dark counts?}
\item {cite more experiments}
\item {Density plots out to larger $|k|$ to show descent into noise floor? How far could we push it until we hit dark counts?}
\item {txy-to-vel vs linear error?}
\item {CHECK THIS 3.}
\item {Clarify \& graph of dark count statistics}
\item {Worst-case bias of -1 counts}
\end{itemize}
In \verb|{40_conclusion.tex}|:
\begin{itemize}
\item {Actually write a conclusion i.e. summary and outlook for the various themes and the field in general.}
\end{itemize}
